---
title: "Stat Eco HW 11 RDA & Variance Decomposition"
author: "Kuei Yueh Ko"
date: "2017年5月11日"
output: html_document
---

# Set Environment
import needed package
```{r}
library(ComplexHeatmap)
library(knitr)
library(dplyr)
library(tidyr)
library(vegan)
```

set directories
```{r}
workdir <- "C:\\Users\\clint\\Documents\\GitHub\\BlogDataAnalysis"
setwd(workdir)

filePathData <- file.path(
    workdir,
    "Data/StatCompEcology/Copepod")
```

# Import data
Read in the copepod data recorded in different stations and preprocess the data to find the dominant species. The final matrix is teh frequency matrix of dominant species in different stations 

We import two dataset:  

1. cop_density.txt
    + copepod total density (number/m^3) recorded in each station
2. copepod_composition.txt
    + frequency of each copepod specie recorded in each station.
3. copepodSPlist.txt
    + Species names along the rows of data copepod_composition.txt

**Note: Definition of dominant species:**

* species >=2% of total composition in any cruise-station

```{r}
# intialize a list to store data
datCPOD <- list()

# read tables: number of copepod in each station
# 34 stations
dat <- read.table(file.path(
    filePathData, "cop_density.txt"), 
    header=T)
datCPOD$Density <- dat

# read tables: freq of copepod in each station
# 34 stations x 181 species
dat <- read.table(file.path(
    filePathData, "copepod_composition.txt"), 
    header=T)
datCPOD$Compose <- dat

# read tables: species name
# 181 species
dat <- read.table(file.path(
    filePathData, "copepodSPlist.txt"),
    sep="\t", header=F)
datCPOD$SP_Total <- as.character(dat$V1)

# Search for dominant species
dat <- datCPOD$Compose
temp <- apply(dat >= 2, 1, sum) 
datCPOD$SP_Dominant <- which(temp > 0)

# Extract the composition of dominant species
dat <- datCPOD$Compose
idx <- datCPOD$SP_Dominant 
idn <- datCPOD$SP_Total
dat <- dat[idx,]
rownames(dat) <- idn[idx]

# composition of dominant species
# Note: transform the matrix since 
# we are clustering on stations
datCPOD$CompDmt <- t(dat)
datCPOD$CompDmtScale <- scale(t(dat)) # scales the columns of a numeric matrix
#head(datCPOD$CompDmtScale)
```

Visualize the data after preprocess (find the dominant species and scale the result data matrix)
```{r}
# Original data matrix of dominant species in different stations
dat <- datCPOD$CompDmt
ht1 <- Heatmap(
    dat, column_title = "Before scaled...",
    row_names_gp = gpar(fontsize = 10),
    column_names_max_height=unit(15, "cm"),
    column_names_gp = gpar(fontsize = 10),
    cluster_rows=F, cluster_columns=F)

# Scaled data matrix of dominant species in different stations
dat <- datCPOD$CompDmtScale
ht2 <- Heatmap(
    dat, column_title = "After scaled...",
    row_names_gp = gpar(fontsize = 10),
    column_names_max_height=unit(15, "cm"),
    column_names_gp = gpar(fontsize = 10),
    cluster_rows=F, cluster_columns=F)

# plot
draw(ht1 + ht2, 
     column_title = "Compare the data before/after scaled",
     column_title_side = "bottom")
```

import environmental data
```{r}
datENV <- read.table(file.path(
    filePathData,
    "enviANDdensity.txt"),
    header=T,
    stringsAsFactors = FALSE)
```

# Redundant Analysis (RDA)
Perform RDA analysis   
    - Constrained:   Rank=11  
    - Unconstrained: Rank=22  

```{r}
dat1 <- datCPOD$CompDmtScale
dat2 <- datENV[,-1]
resRDA <- rda(dat1~., dat2)
print(resRDA)
```

WA & LC scores
```{r}
# set plot
par(mfrow=c(1,2))

# plot
plot(resRDA, display="lc", main="LC scores")
plot(resRDA, display="wa", main="WA scores")

# reset plot
par(mfrow=c(1,1))
```

If environmental variables correlate with species data, WA and LC 
scores should resemble each other
```{r}
spenvcor(resRDA)
```

It turns out that the correlation is quite high. This can be observed in the plot of WA and LC scores (similar distribution of stations)
```{r}
plot(resRDA, dis=c("wa","lc"))
#      RDA1      RDA2      RDA3      RDA4
# 0.9652628 0.8890507 0.8841969 0.8668161
    
```

We can plot the correlation of environment variables and species variables
```{r}
scoreLC <- scores(resRDA, display="lc", choice=c(1:5))
scoreWA <- scores(resRDA, display="wa", choice=c(1:5))
mat <- cor(scoreLC, scoreWA)
Heatmap(
    mat,
    row_names_gp = gpar(fontsize = 15),
    column_names_max_height=unit(15, "cm"),
    column_names_gp = gpar(fontsize = 15),
    cluster_rows=F, cluster_columns=F)
```

biplot using scaling=3
```{r}
ordiplot(resRDA, type="text", scaling=3)
```

# CCA analysis
```{r}
dat1 <- datCPOD$CompDmt
dat2 <- datENV[,-1]
resCCA <- cca(dat1~., dat2)
print(resCCA)
```

```{r}
plot(resCCA, scaling=3,
     xlim=c(-4,4),
     ylim=c(-2,2))
```


# Variance Decomposition
```{r}
dat <- datCPOD$CompDmtScale
datEnv     <- datENV[, -1]
```

```{r}
# set factors
idn <- names(datENV)
idnPhys <- idn[2:10]
idnBio  <- idn[11:12] 

# biological factors
formP <- formula(paste("dat ~", paste(idnPhys, collapse = " + ")))

# physical factors
formB <- formula(paste("dat ~", paste(idnBio , collapse = " + ")))

# biological factors condition on physical factors
formPB <- formula(
    paste("dat ~", 
              paste(idnPhys, collapse = " + "),
          "+ Condition(", 
              paste(idnBio, collapse = " + "), 
          ")"))

# physical factors condition on biological factors
formBP <- formula(
    paste("dat ~", 
              paste(idnBio, collapse = " + "),
          "+ Condition(", 
              paste(idnPhys, collapse = " + "), 
          ")"))
```

perform RDA using different type of variance decomposition
```{r}
resRDA <- list()
resRDA$Tot  <- rda(dat~., datEnv)
resRDA$Phys <- rda(formP, datEnv)
resRDA$Bio  <- rda(formB, datEnv)
resRDA$PB   <- rda(formPB, datEnv)
resRDA$BP   <- rda(formBP, datEnv)
```

plot the results
```{r}
ordiplot(resRDA$Tot,  type="text", scaling=3, main="Total")
ordiplot(resRDA$Phys, type="text", scaling=3, 
         main="Physical Factors")
ordiplot(resRDA$PB,   type="text", scaling=3, 
         main="Physical Factors Conditional on Biological Factors")
ordiplot(resRDA$Bio,  type="text", scaling=3, 
         main="Biological Factors")
ordiplot(resRDA$BP,   type="text", scaling=3, 
         main="Biological Factors Conditional on Physical Factors")
```

