---
title: "Stat Eco HW 11 RDA & Variance Decomposition"
author: "Kuei Yueh Ko"
date: "2017年5月11日"
output: html_document
---

# Set Environment
import needed package
```{r}
library(ComplexHeatmap)
library(knitr)
```

set directories
```{r}
workdir <- "C:\\Users\\clint\\Documents\\GitHub\\BlogDataAnalysis"
setwd(workdir)

filePathData <- file.path(
    workdir,
    "Data/StatCompEcology/Copepod")
```

# Import data
Read in the copepod data recorded in different stations and preprocess the data to find the dominant species. The final matrix is teh frequency matrix of dominant species in different stations 

We import two dataset:  

1. cop_density.txt
    + copepod total density (number/m^3) recorded in each station
2. copepod_composition.txt
    + frequency of each copepod specie recorded in each station.
3. copepodSPlist.txt
    + Species names along the rows of data copepod_composition.txt

**Note: Definition of dominant species:**

* species >=2% of total composition in any cruise-station

```{r}
# intialize a list to store data
datCPOD <- list()

# read tables: number of copepod in each station
# 34 stations
dat <- read.table(file.path(
    filePathData, "cop_density.txt"), 
    header=T)
datCPOD$Density <- dat

# read tables: freq of copepod in each station
# 34 stations x 181 species
dat <- read.table(file.path(
    filePathData, "copepod_composition.txt"), 
    header=T)
datCPOD$Compose <- dat

# read tables: species name
# 181 species
dat <- read.table(file.path(
    filePathData, "copepodSPlist.txt"),
    sep="\t", header=F)
datCPOD$SP_Total <- as.character(dat$V1)

# Search for dominant species
dat <- datCPOD$Compose
temp <- apply(dat >= 2, 1, sum) 
datCPOD$SP_Dominant <- which(temp > 0)

# Extract the composition of dominant species
dat <- datCPOD$Compose
idx <- datCPOD$SP_Dominant 
idn <- datCPOD$SP_Total
dat <- dat[idx,]
rownames(dat) <- idn[idx]

# composition of dominant species
# Note: transform the matrix since 
# we are clustering on stations
datCPOD$CompDmt <- t(dat)
datCPOD$CompDmtScale <- scale(t(dat)) # scales the columns of a numeric matrix
#head(datCPOD$CompDmtScale)
```

Visualize the data after preprocess (find the dominant species and scale the result data matrix)
```{r}
# Original data matrix of dominant species in different stations
dat <- datCPOD$CompDmt
ht1 <- Heatmap(
    dat, column_title = "Before scaled...",
    row_names_gp = gpar(fontsize = 10),
    column_names_max_height=unit(15, "cm"),
    column_names_gp = gpar(fontsize = 10),
    cluster_rows=F, cluster_columns=F)

# Scaled data matrix of dominant species in different stations
dat <- datCPOD$CompDmtScale
ht2 <- Heatmap(
    dat, column_title = "After scaled...",
    row_names_gp = gpar(fontsize = 10),
    column_names_max_height=unit(15, "cm"),
    column_names_gp = gpar(fontsize = 10),
    cluster_rows=F, cluster_columns=F)

# plot
draw(ht1 + ht2, 
     column_title = "Compare the data before/after scaled",
     column_title_side = "bottom")
```

import environmental data
```{r}
datENV <- read.table(file.path(filePathData,"enviANDdensity.txt"))
```

